Including EW and mixed QCD-EW corrections into PDF fits

### 2020-07-20 paper/sec-pineappl.tex
Revise PineAPPL section

In this paper we present a new library called \textsc{PineAPPL}, which stores phase-space weights of a Monte Carlo (MC) integration of a fixed-order calculation independently from the chosen PDFs.
This separates the task of producing predictions for observables into two steps: 1) the generation of the \emph{grids}, which is how the generated files are called, and 2) the convolution of those grids with a set of PDFs.
The advantage of this method is that the time-consuming step 1) has to be done only once and, therefore, allows for very quick convolutions of a given grid with multiple PDF sets.

@@

If chosen dynamically the Ellis--Sexton scale $Q^2$ depends on the phase space, but we assume the fractions $\xi_\mathrm{R}$ and $\xi_\mathrm{F}$ to be constants of phase space in any case.
This allows variations around the central scale choice $\xi_\mathrm{R} = \xi_\mathrm{F} = 1$, but not changing the scale arbitrarily.
The terms with powers $m > 0$ and $n > 0$ vanish for the central scale choice and are only required for variations of the factorisation and renormalisation scales.
To estimate the perturbative QCD uncertainty---no EW uncertainty is covered by this method---one typically uses a 7-point scale variation, which evaluates the cross section using the following values,
\begin{equation}
(\xi_\mathrm{R}, \xi_\mathrm{F}) \in \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr) \right\} \text{.}
\end{equation}
The (asymmetric) uncertainties are then given as the minimum and maximum value (the envelope), measured from the central value $(1, 1)$.
As is clear from eq.~\eqref{eq:expansion}, the EW coupling $\alpha$ is assumed not to be a dynamically varying coupling, but instead a constant over phase space.
This, however, includes the most common choices of the coupling, which are (not necessarily in this order), $\alpha (0)$, $\alpha (M_\mathrm{Z})$, and $\alpha_{G_\mu}$.
The task that \textsc{PineAPPL} solves can now be described: Approximately reconstruct the functions
\begin{equation}
w_{ab}^{(k,l,m,n,o)} \left( x_1, x_2, Q^2 \right)
\end{equation}
from a set of $N$ function evaluations for specific momentum fractions, scales, and values of the observable:
\begin{equation}
\left\{ x_1^{(i)}, x_2^{(i)}, Q^2_i, \mathcal{O}_i \right\}_{i=1}^N \text{,}
\end{equation}
given by the MC integrator.
Using eq.~\eqref{eq:expansion} and
\begin{multline}
\frac{\mathrm{d} \sigma}{\mathrm{d} \mathcal{O}} (\mathcal{O}, \xi_\mathrm{R}, \xi_\mathrm{R}) \\
= \sum_{a,b} \int_0^1 \mathrm{d} x_1 \int_0^1 \mathrm{d} x_2 \int_{Q^2_\mathrm{min}}^{Q^2_\mathrm{max}} \mathrm{d} Q^2 \, f_a (x_1, \xi_\mathrm{F}^2 Q^2) f_b (x_2, \xi_\mathrm{F}^2 Q^2) \sigma_{ab} (x_1, x_2, Q^2, \xi_\mathrm{R}, \xi_\mathrm{R}) \text{,}
\label{eq:pineappl-convolution}
\end{multline}
\textsc{PineAPPL} can then quickly calculate hadronic cross sections for arbitrarily many PDF sets and perform scale variations.

@@

\subsection{Grid representation and accuracy}
\label{sec:grid-representation}

@@

\subsubsection{Interpolation grids}

A different choice partitions the $(x_1, x_2, Q^2)$ space,
\begin{equation}
H = [x_\mathrm{min},x_\mathrm{max}]^2 \times [Q^2_\mathrm{min}, Q^2_\mathrm{max}] \ni (x_1, x_2, Q^2)
\end{equation}
along each axis into a small numbers of bins and to insert the phase-space weights $w$ into the corresponding discrete bin.
This method is straightforward, but given a finite number of bins this usually yields a bad approximation for the cross section.
Increasing the number of bins improves the accuracy, but it also increases the space requirements of the grid.
Interpolation methods offer more precision, while keeping the number of bins constants.

We use the same Lagrange-interpolation grid presented in Ref.~\cite{Carli:2010rw} with the parameters published in Ref.~\cite{Bertone:2014zva}, which give decent precision (see section~\ref{sec:results}).
The method presented in the previous references first maps $(x_1, x_2, Q^2) \mapsto (y_1, y_2, \tau)$, with the maps defined as
\begin{equation}
y(x) = 5 (1-x) - \log x \text{,} \qquad \tau (Q^2) = \log \log \frac{Q^2}{(\SI{0.25}{\giga\electronvolt})^2} \text{,}
\end{equation}
and then feeds the variables $(y_1, y_2, \tau)$ into the 3-dimensional Lagrange-interpolation grid with 50 bins in each $y$ direction and 30 bins in $\tau$ direction.
The interpolation order is 3 for each variable, and only the subspace $[\num{2e7},1] \times [\num{2e-7},1] \times [100,\num{e6}] \subset H$\footnote{In Ref.~\cite{Bertone:2014zva} the upper limit for $Q$ is given as \SI{3162}{\giga\electronvolt} (which corresponds to $Q_\mathrm{max}^2 \approx \SI{e7}{\giga\electronvolt}$), but in the code we found the value $Q_\mathrm{max}^2 = \SI{e6}{\giga\electronvolt\squared}$.} is mapped.
For the convolution of a grid with a PDF set also the inverse functions are needed, which are
\begin{equation}
x(y) = \frac{1}{5} \operatorname{W}_0 (5 \exp (5-x)) \text{,} \qquad Q^2 (\tau) = (\SI{0.25}{\giga\electronvolt})^2 \exp (\exp (\tau)) \text{,}
\end{equation}
where $\operatorname{W}_0 (x)$ is (the principle branch of) the Lambert W function or product logarithm, which satisfies the relation $\operatorname{W} (x) \exp (\operatorname{W} (x)) = x$.

@@

\begin{figure}
\centering
\parbox{0.6\textwidth}{\includegraphics{figures/grid-insertion}}
\begin{tabular}{ll}
\toprule
$u_1$/$u_2$ & $x_1$/$x_2$ \\
\midrule
0 & \num{1.00e0} \\
1 & \num{6.36e-1} \\
2 & \num{3.20e-1} \\
3 & \num{9.96e-2} \\
4 & \num{1.57e-2} \\
5 & \num{1.74e-3} \\
6 & \num{1.81e-4} \\
7 & \num{1.87e-5} \\
8 & \num{1.93e-6} \\
9 & \num{2.00e-7} \\
\bottomrule
\end{tabular}
\caption{Example for a 2-dimensional $10 \times 10$ grid, which is being filled at the location marked with the small black square, at $(5.8,4.8)$.
Each side of the grey square starting at $k_i = 4$ and $k_j = 3$ has a length of $N_y + 1 = 4$.
This square marks the grid values that are being updated using eq.~\eqref{eq:interpolation}.
The table on the right-hand side marks gives the parton-momentum fractions for each grid point.
Note that around $u=3$ the values are linearly distributed, then logarithmically.}
\label{fig:grid}
\end{figure}

>>>>

In this paper we present a new library called \textsc{PineAPPL}, which stores phase-space weights of a Monte Carlo (MC) integration of a fixed-order calculation independently from the chosen PDFs.
This separates the task of producing predictions for observables into two steps: 1) the generation of the \emph{grids}, which is how the generated files are called, and 2) the convolution of those grids with a set of PDFs.
The advantage of this method is that the time-consuming step 1) has to be done only once and, therefore, allows in step 2) for very quick convolutions of a given grid with multiple PDF sets.

@@

If chosen dynamically the Ellis--Sexton scale $Q^2$ depends on the phase space, but we assume the fractions $\xi_\mathrm{R}$ and $\xi_\mathrm{F}$ to be constants of phase space in any case.
This allows variations around the central scale choice $\xi_\mathrm{R} = \xi_\mathrm{F} = 1$, but not changing the scale arbitrarily.
The terms with powers $m > 0$ and $n > 0$ vanish for the central scale choice and are only required for variations of the factorisation and renormalisation scales.
To estimate the perturbative QCD uncertainty---no EW uncertainty is covered by this method---one typically uses a 7-point scale variation, which evaluates the cross section using the following values,
\begin{equation}
(\xi_\mathrm{R}, \xi_\mathrm{F}) \in \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr) \right\} \text{.}
\end{equation}
The (asymmetric) uncertainties are then given as the minimum and maximum value (the envelope), measured from the central value $(1, 1)$.
As is clear from eq.~\eqref{eq:expansion}, the EW coupling $\alpha$ is assumed not to be a dynamically varying coupling, but instead a constant over phase space.
This, however, includes the most common choices of the coupling, which are (not necessarily in this order), $\alpha (0)$, $\alpha (M_\mathrm{Z})$, and $\alpha_{G_\mu}$.
The task that \textsc{PineAPPL} solves can now be described: Approximately reconstruct the functions
\begin{equation}
w_{ab}^{(k,l,m,n,o)} \left( x_1, x_2, Q^2 \right)
\label{eq:weight-map}
\end{equation}
from a set of $N$ function evaluations for specific momentum fractions, scales, and values of the observable:
\begin{equation}
\left\{ x_1^{(i)}, x_2^{(i)}, Q^2_i, \mathcal{O}_i \right\}_{i=1}^N \text{,} \label{eq:phase-space-weights}
\end{equation}
given by the MC integrator.

Using eq.~\eqref{eq:expansion} and
\begin{multline}
\frac{\mathrm{d} \sigma}{\mathrm{d} \mathcal{O}} (\mathcal{O}, \xi_\mathrm{R}, \xi_\mathrm{R}) \\
= \sum_{a,b} \int_0^1 \mathrm{d} x_1 \int_0^1 \mathrm{d} x_2 \int_{Q^2_\mathrm{min}}^{Q^2_\mathrm{max}} \mathrm{d} Q^2 \, f_a (x_1, \xi_\mathrm{F}^2 Q^2) f_b (x_2, \xi_\mathrm{F}^2 Q^2) \sigma_{ab} (x_1, x_2, Q^2, \xi_\mathrm{R}, \xi_\mathrm{R}) \text{,}
\label{eq:pineappl-convolution}
\end{multline}
\textsc{PineAPPL} can then quickly calculate hadronic cross sections for arbitrarily many PDF sets and perform scale variations.

@@

\subsection{Grid representations}
\label{sec:grid-representation}

@@

\subsubsection{Lagrange-interpolation grid}

A different strategy is to partition the $(x_1, x_2, Q^2)$ space,
\begin{equation}
H = [x_\mathrm{min},x_\mathrm{max}]^2 \times [Q^2_\mathrm{min}, Q^2_\mathrm{max}] \ni (x_1, x_2, Q^2)
\end{equation}
along each axis into a small numbers of bins and to insert the phase-space weights $w$ into the corresponding discrete bin.
Using the bin centres and their values one already has a straightforward implementation of eq.~\eqref{eq:weight-map}, but given a finite number of bins it usually yields an insufficient approximation for the cross section.
Increasing the number of bins improves the accuracy, but it also increases the space requirements.
Interpolation methods offer more accuracy with the same number of bins.

We use the same \enquote{Lagrange-interpolation grid} presented in ref.~\cite{Carli:2010rw} with the parameters published in ref.~\cite{Bertone:2014zva}, which give sufficient precision (see section~\ref{sec:results}).

This method first maps $(x_1, x_2, Q^2) \mapsto (y_1, y_2, \tau)$, with
\begin{equation}
y(x) = 5 (1-x) - \log x \text{,} \qquad \tau (Q^2) = \log \log \frac{Q^2}{(\SI{0.25}{\giga\electronvolt})^2} \text{.}
\label{eq:maps}
\end{equation}
The function $y(x)$ maps events with large $x$ effectively linearly and small $x$ effectively logarithmically onto $y$.
This reflects our knowledge of PDFs, which behave differently in those regions, and thereby increases the accuracy of the interpolation.
For the convolution of a grid with a PDF set also the inverse functions are needed, which are
\begin{equation}
x(y) = \frac{1}{5} \operatorname{W}_0 (5 \exp (5-y)) \text{,} \qquad Q^2 (\tau) = (\SI{0.25}{\giga\electronvolt})^2 \exp (\exp (\tau)) \text{,}
\end{equation}
where $\operatorname{W}_0 (x)$ is (the principle branch of) the Lambert W function or product logarithm, which satisfies the relation $\operatorname{W} (x) \exp (\operatorname{W} (x)) = x$.

@@

\begin{figure}
\centering
\parbox{0.6\textwidth}{\includegraphics{figures/grid-insertion}}
\begin{tabular}{ll}
\toprule
$u_1$/$u_2$ & $x_1$/$x_2$ \\
\midrule
0 & \num{1.00e0} \\
1 & \num{6.36e-1} \\
2 & \num{3.20e-1} \\
3 & \num{9.96e-2} \\
4 & \num{1.57e-2} \\
5 & \num{1.74e-3} \\
6 & \num{1.81e-4} \\
7 & \num{1.87e-5} \\
8 & \num{1.93e-6} \\
9 & \num{2.00e-7} \\
\bottomrule
\end{tabular}
\caption{Example for a 2-dimensional $10 \times 10$ grid, which is being filled at the location marked with the small black square at $(5.8,4.8)$.
Each side of the grey square starting at $k_i = 4$ and $k_j = 3$ has a length of $N_y + 1 = 4$.
This square marks the grid values (grey dots) that are being updated using eq.~\eqref{eq:interpolation}.
The table on the right-hand side gives the parton-momentum fractions for each grid point according to eq.~\eqref{eq:maps}.
Note that for $u \in [0, 3]$ the values are roughly linearly distributed, then logarithmically.}
\label{fig:grid}
\end{figure}
